Practical Optional Types for Clojure

### 2015-02-24 formal-model.tex
experience

Figures ~\ref{main:figure:typingrules} and~\ref{main:figure:standardopsem}
contain the typing rules and operational semantics.
They compensate for the extra logically false value \nil{}, otherwise they are standard.

@@

\begin{figure}
  \footnotesize
$$
\begin{array}{lrlr}
  \expd{}, \e{} &::=& \x{}
                      \alt \v{} 
                      \alt {\comb {\e{}} {\e{}}} 
                      \alt {\abs {\x{}} {\t{}} {\e{}}} &\mbox{Expressions} \\
                      &\alt& {\ifexp {\e{}} {\e{}} {\e{}}}
                      %\alt {\trdiff{\doexp {\e{}} {\e{}}}}
                      \alt {\letexp {\x{}} {\e{}} {\e{}}}\\
                      %\alt {\errorvalv{}}
  \v{} &::=&          \singletonmeta{}
                      \alt {\const{}}
                      \alt {\closure {\openv{}} {\abs {\x{}} {\t{}} {\e{}}}}
                &\mbox{Values} \\
                \constantssyntax{}\\
  \s{}, \t{}    &::=& \Top 
                      \alt {\Unionsplice {\overrightarrow{\t{}}}}
                      \alt
                      {\ArrowOne {\x{}} {\t{}}
                                   {\t{}}
                                   {\filterset {\prop{}} {\prop{}}}
                                   {\object{}}}
                &\mbox{Types} \\
                      &\alt& {\Value \singletonmeta{}} 
                      \alt \trdiff{\class{}}\\
  \singletonallsyntax{}
                \\ \\
  \occurrencetypingsyntax{}\\
  \openvsyntax{}
  %\\
  %\classliteralallsyntax{}
\end{array}
$$
\caption{Syntax of Terms, Types, Propositions and Objects}
\label{main:figure:termsyntax}
\end{figure}

@@

\begin{figure}
  \footnotesize
  \begin{mathpar}
%    {\BLocal}
%
%    %{\BDo}
%
%    {\BLet}
%
%    \BVal{}
%
    \BIfTrue{}

%    \BIfFalse{}
%
%    \BAbs{}
%
%    \BBetaClosure{}
%
%    \BDelta{}
  \end{mathpar}
  \caption{Operational Semantics}
  \label{main:figure:standardopsem}
\end{figure}

@@

We extend our model with sequencing expressions and errors, where {\errorvalv{}}
is equivalent to a Clojure throw expression.

@@

@@

\begin{figure}
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \hmapexpressionsyntax{}
    &\mbox{Expressions} \\
    \v{} &::=& \ldots \alt {\emptymap{}}
    &\mbox{Values} \\
    \t{} &::=& \ldots \alt {\HMapgeneric {\mandatory{}} {\absent{}}}
    &\mbox{Types} \\
    \auxhmapsyntax{}\\
  \end{altgrammar}
  $$
  \caption{HMap Terms and Types}
  \label{main:figure:hmapsyntax}
\end{figure}

It is common in Clojure to use plain immutable hash maps to represent ad-hoc
objects with a known set of keyword entries. We use \HMapliteral{}
types (figure~\ref{main:figure:hmapsyntax}) to track known present, known absent and unknown keyword entries.

The metavariable \completenessmeta{} tracks whether a HMap is fully ({\complete{}}) or partially ({\partial{}})
specified. \mandatory{} specifies a set of known present keyword entries, \absent{} tracks known absent keyword
entries and we write \HMapcwithabsent{\mandatory{}}{\absent{}} as \HMapc{\mandatory{}} as a fully specified
HMap can assume all entries not in {\mandatory{}} as absent. We disallow keys to be both present and absent.

\getliteral{} and \assocliteral{} lookup and extend immutable maps respectively.
To simplify presentation, we only allow the empty map literal and
restrict lookup and extension to keyword keys. The metavariable \mapval{}
ranges over the runtime value of maps {\curlymapvaloverright{\k{}}{\v{}}}.

Figure~\ref{main:figure:hmapsem} shows the semantics for HMaps, where B-Get returns the associated
entry if it is present and B-GetMissing returns \nil{} otherwise. B-Assoc returns a new extended
map.

\begin{figure*}
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \reflectiveexpsyntax{}
            \alt  \localhintsyntax{} \alt \lethintsyntax{}
    &\mbox{Expressions}
  \end{altgrammar}
  $$
  \caption{Java Interoperability Terms and Types}
  \label{main:figure:reflectivesyntax}
\end{figure*}

We extend the external language with Java field lookups and calls to
methods and constructors (figure~\ref{main:figure:nonreflectivesyntax}).
There are no typing or evaluation rules for these syntaxes because
Typed Clojure must be convinced that Java interactions are \emph{non-reflective},
ie., are unambiguously resolved at compile-time. Programmers use
type hinted expressions (figure~\ref{main:figure:nonreflectivesyntax}) 
to help the compiler eliminate reflection.

\begin{figure}
  \footnotesize
  \begin{mathpar}
    %\RAbs{}
%
%    \RNewElimRefl{}
%
%    \RMethodElimRefl{}
%
    \RFieldElimRefl{}
%
%    \RLet{}
%
    \RLetHint{}
  \end{mathpar}
\caption{Reflection Elimination (select rules, figure~\ref{appendix:figure:rewrite} for full rules)}
\label{main:figure:rewrite}
\end{figure}

core.typed uses tools.analyzer to obtain a rich abstract syntax tree that includes resolved
reflection, however to tell the full story we include a rewrite relation

to eliminate reflection when sufficient type hints are provided.

To help infer type hints

gives the type hint \tatype{} of expression \e{} in environemtn \taenv{}.

$$
\TALetHint{}
$$

{\footnotesize
\input{constant}

>>>>

Figure ~\ref{main:figure:typingrules} contains the standard typing rules.
The operational semantics are standard, except \nil{} is logically false:

\begin{mathpar}
    \BIfTrue{}

    \BIfFalse{}
\end{mathpar}

@@

\begin{figure}
  \footnotesize
$$
\begin{array}{lrll}
  \expd{}, \e{} &::=& \x{}
                      \alt \v{} 
                      \alt {\comb {\e{}} {\e{}}} 
                      \alt {\abs {\x{}} {\t{}} {\e{}}} &\mbox{Expressions} \\
                      &\alt& {\ifexp {\e{}} {\e{}} {\e{}}}
                      %\alt {\trdiff{\doexp {\e{}} {\e{}}}}
                      \alt {\letexp {\x{}} {\e{}} {\e{}}}\\
                      %\alt {\errorvalv{}}
  \v{} &::=&          \singletonmeta{}
                      \alt {\const{}}
                      \alt {\closure {\openv{}} {\abs {\x{}} {\t{}} {\e{}}}}
                &\mbox{Values} \\
                \constantssyntax{}\\
  \s{}, \t{}    &::=& \Top 
                      \alt {\Unionsplice {\overrightarrow{\t{}}}}
                      \alt
                      {\ArrowOne {\x{}} {\t{}}
                                   {\t{}}
                                   {\filterset {\prop{}} {\prop{}}}
                                   {\object{}}}
                &\mbox{Types} \\
                      &\alt& {\Value \singletonmeta{}} 
                      \alt \trdiff{\class{}}\\
  \singletonallsyntax{}
                \\ \\
  \occurrencetypingsyntax{}\\
  \propenvsyntax{}\\
  \openvsyntax{}
  %\\
  %\classliteralallsyntax{}
\end{array}
$$
\caption{Syntax of Terms, Types, Propositions and Objects}
\label{main:figure:termsyntax}
\end{figure}

@@

%\begin{figure}
%  \footnotesize
%  \begin{mathpar}
%    {\BLocal}
%
%    %{\BDo}
%
%    {\BLet}
%
%    \BVal{}
%
    %\BIfTrue{}

%    \BIfFalse{}
%
%    \BAbs{}
%
%    \BBetaClosure{}
%
%    \BDelta{}
%  \end{mathpar}
%  \caption{Operational Semantics}
%  \label{main:figure:standardopsem}
%\end{figure}

@@

We extend our model with sequencing expressions and errors, where {\errorvalv{}}
is equivalent to Clojure's throw.

@@

\begin{figure}
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \hmapexpressionsyntax{}
    &\mbox{Expressions} \\
    \v{} &::=& \ldots \alt {\emptymap{}}
    &\mbox{Values} \\
    \t{} &::=& \ldots \alt {\HMapgeneric {\mandatory{}} {\absent{}}}
    &\mbox{Types} \\
    \auxhmapsyntax{}\\
    \pesyntax{}   &::=& \ldots \alt {\keype{\k{}}}
                  &\mbox{Path Elements}
  \end{altgrammar}
  $$
  \caption{HMap Terms and Types}
  \label{main:figure:hmapsyntax}
\end{figure}

We now present \HMapliteral{} types in detail.
Figure~\ref{main:figure:hmapsyntax} extends the syntax for terms and types.
\clj{(HMap :mandatory p :absent-keys a)} is written \HMapgeneric{\mandatory{}}{\absent{}}.
The metavariable \completenessmeta{} tracks whether a HMap is fully ({\complete{}}) or partially ({\partial{}})
specified. \mandatory{} specifies a set of known present keyword entries, \absent{} tracks known absent keyword
entries and we abbreviate \HMapcwithabsent{\mandatory{}}{\absent{}} as \HMapc{\mandatory{}}. Keys cannot be both present and absent.

\getliteral{} and \assocliteral{} lookup and extend immutable maps respectively.
To simplify presentation, we only provide syntax for the empty map literal and
restrict lookup and extension to keyword keys. The metavariable \mapval{}
ranges over the runtime value of maps {\curlymapvaloverright{\k{}}{\v{}}}.

We introduce our first path element {\keype{\k{}}}, the symbolic representation
of a \getliteral{} expression on keyword \k{} where the sentinel value for
a missing entry is \nil{}.

Figure~\ref{main:figure:hmapsem} shows the semantics for HMaps. B-Get returns the associated
value if the entry is present otherwise  B-GetMissing returns \nil{}. B-Assoc returns a new extended
map.

  $$
  \begin{altgrammar}
    \e{} &::=& \ldots   \localhintsyntax{} \alt \lethintsyntax{} &\mbox{Expressions}\\
            &\alt& \reflectiveexpsyntax{} 
  \end{altgrammar}
  $$

We extend the external language with Java field lookups and calls to
methods and constructors.
These come without typing or evaluation rules:
Typed Clojure requires the Clojure compiler to statically resolve all Java interactions
in typed code.
Type hinted expressions such as \localhintsyntax{} 
to help the compiler eliminate reflection.

\begin{figure}
  \footnotesize
  \begin{mathpar}
    %\RAbs{}
%
%    \RNewElimRefl{}
%
%    \RMethodElimRefl{}
%
    \RFieldElimRefl{}
%
%    \RLet{}
%

    \RLetHint{}
  \end{mathpar}
\caption{Reflection Elimination %(select rules, figure~\ref{appendix:figure:rewrite} for full rules)
}
\label{main:figure:rewrite}
\end{figure}

rewrites expression \e{} to a possibly-less reflective expression
\ep{} with respect to type hint environment
\taenv{} and Java class table \ct{}.

infers the (possibly-unknown) type hint \tatype{} of expression \e{} in environment \taenv{}.

%$$
%\TALetHint{}
%$$

\subsection{\classconst{}}

The \classconst{} function is modeled as a constant which returns
an object's \Class{}, or \nil{} if provided \nil{}.

\begin{mathpar}
\constanttypefigure{}
\end{mathpar}

Our second path element \classpe{} is the symbolic representation
of a call to \classconst{}.

$$
\begin{array}{lrlr}
  \pesyntax{}   &::=& \ldots \alt {\classpe{}}
                &\mbox{Path Elements}
\end{array}
$$

@@

{\footnotesize
\constantsemfigure{}
